---
title: "Data preparation"
output:
  pdf_document: default
---

# Instructions

- You only need to submit the .Rmd of this file, not a PDF.

- You should __comment__ your code clearly to show what you've done to prepare the data.

- The purpose of this file is to use the data in the `data-raw` folder to create the data you will use in the report. The data you will use in the report should be saved in the `data` folder. It is good professional practice to make sure you're never directly modifying your raw data, but instead creating new datasets based on merges/manipulations that you need to reuse.

- Make sure you've taken a look at the hints for the web scraping and census API. 

- You may find the `write_rds()` function from the `readr` package helpful (it is loaded as part of the `tidyverse`).

- You do not need to keep the structure below.

# Set up

```{r, libraries}
# Set up any libraries you need
library(tidyverse)
library(polite)
library(rvest)
library(haven)
library(lme4)
library(eeptools)
library(knitr)
library(sjPlot)
library(cancensus)
library(janitor)
```

# Loading client data

```{r}
customer <- read_rds("data-raw/customer.Rds")
device <- read_rds("data-raw/device.Rds")
cust_sleep <- read_rds("data-raw/cust_sleep.Rds")
cust_dev <- read_rds("data-raw/cust_dev.Rds")
break_glass <- read_rds("data-raw/break_glass_in_case_of_emergency.Rds")
```

## Web scraping industry data

```{r}
url <- "https://fitnesstrackerinfohub.netlify.app/"
target <- bow(url,
              user_agent = "alexandertran.tran@mail.utoronto.ca for STA303/1002 project",
              force = TRUE)
target

html <- scrape(target)

device_data <- html %>% 
  html_elements("table") %>% 
  html_table() %>% 
  pluck(1) # added, in case you're getting a list formats

```
## Postcode

```{r}
postcode <- read_sav("pccfNat_fccpNat_082021sav.sav", col_select = c("PC", "CSDuid"))

```

## Census API

```{r}
options(cancensus.api_key = "CensusMapper_e044495035736957fbd42a934af93f49",
        cancensus.cache_path = "cache") # this sets a folder for your cache


# get all regions as at the 2016 Census (2020 not up yet)
regions <- list_census_regions(dataset = "CA16")

regions_filtered <-  regions %>% 
  filter(level == "CSD") %>% # Figure out what CSD means in Census data
  as_census_region_list()



# This can take a while
# We want to get household median income
census_data_csd <- get_census(dataset='CA16', regions = regions_filtered,
                          vectors=c("v_CA16_2397"), 
                          level='CSD', geo_format = "sf")


#Simplify to only needed variables

median_income <- census_data_csd %>% 
  as_tibble() %>% 
  select(CSDuid = GeoUID, contains("median"), Population) %>% 
  mutate(CSDuid = parse_number(CSDuid)) %>% 
  rename(hhld_median_inc = 2)
```
## Data cleaning and processing

```{r}
# Rename the column of device_data so that we can combine it with device
device_data_adj <- clean_names(device_data) %>% mutate(released=as.Date(released))

# Set the released column in device into date type 
device_adj <- clean_names(device) %>% mutate(released=as.Date(released))

# Join device_adj and device_data_adj
device_full <- device_adj %>% right_join(device_data_adj, by=c("line", "device_name", "released"))

cust_dev_joined <- cust_sleep %>% left_join(customer, by="cust_id") %>% left_join(cust_dev, by="cust_id") %>% left_join(device_full, by="dev_id")

# Clean the postcode data and create a new dataframe postcode_adj
postcode_adj <- clean_names(postcode)
postcode_adj <- rename(postcode_adj, postcode=pc)

# Clean the median_income data and create a new dataframe median_income_adj
# Combine median_income_adj and postcode_adj
median_income_adj <- clean_names(median_income)
pc_income <- postcode_adj %>% left_join(median_income_adj, by="cs_duid")

# Combining with postal code and income
cust_dev_joined <- cust_dev_joined %>% left_join(pc_income, by="postcode")

# Cleaning variables of interest
cust_dev_joined <- cust_dev_joined %>% mutate(dob=as.Date(dob))
cust_dev_joined <- cust_dev_joined %>% mutate(age=floor(age_calc(dob, units = "years")))
cust_dev_joined <- cust_dev_joined %>% drop_na(cust_id)
cust_dev_joined$emoji_modifier[is.na(cust_dev_joined$emoji_modifier)] <- "Not Set"
cust_dev_joined$sex[is.na(cust_dev_joined$sex)] <- "Not Set"
```

## Initial Exploratory Data Analysis

```{r}
#Data summaries
#Note that means and variances are calculated as a flag rate of per 60 minutes of sleep
cust_dev_joined %>% group_by(emoji_modifier) %>% summarise(n = n(), mean = mean(flags*60/duration), median = median(flags*60/duration), sd = sd(flags*60/duration), var = var(flags*60/duration))

cust_dev_joined %>% group_by(sex) %>% summarise(n = n(), mean = mean(flags*60/duration), median = median(flags*60/duration), sd = sd(flags*60/duration), var = var(flags*60/duration))

cust_dev_joined %>% mutate(age_bracket = cut(age, breaks = seq(15, 95, by=5))) %>%  group_by(age_bracket) %>% summarise(n = n(), mean = mean(flags*60/duration), median = median(flags*60/duration), sd = sd(flags*60/duration), var = var(flags*60/duration))

cust_dev_joined %>% group_by(device_name) %>% summarise(n = n(), mean = mean(flags*60/duration), median = median(flags*60/duration), sd = sd(flags*60/duration), var = var(flags*60/duration))

cust_dev_joined %>% mutate(income_bracket = cut(hhld_median_inc, breaks = seq(40000, 120000, by=10000))) %>%  group_by(income_bracket) %>% summarise(n = n(), mean = mean(flags*60/duration), median = median(flags*60/duration), sd = sd(flags*60/duration), var = var(flags*60/duration))
```

```{r}
#Visualize flags distribution through histograms
cust_dev_joined %>% ggplot(aes(x = flags)) + geom_histogram(colour="black", fill="white", binwidth=0.5) + labs(title="Distribution of Flags for all Sleep Sessions",x="Flags", y = "Frequency") + theme_minimal()

#By emoji_modifier
cust_dev_joined %>% ggplot(aes(x = flags)) + geom_histogram(colour="black", fill="white", binwidth=0.5) + labs(title="Distribution of Flags by Emoji Modifier",x="Flags", y = "Frequency") + theme_minimal() + facet_wrap(~emoji_modifier, nrow = 3)

#By sex
cust_dev_joined %>% ggplot(aes(x = flags)) + geom_histogram(colour="black", fill="white", binwidth=0.5) + labs(title="Distribution of Flags by Sex",x="Flags", y = "Frequency") +  theme_minimal() + facet_wrap(~sex, nrow = 1)

#By device
cust_dev_joined %>% ggplot(aes(x = flags)) + geom_histogram(colour="black", fill="white", binwidth=0.5) + labs(title="Distribution of Flags By Device",x="Flags", y = "Frequency") + theme_minimal() + facet_wrap(~device_name, nrow = 4)

#By age
cust_dev_joined %>% mutate(age_bracket = cut(age, breaks = seq(15, 95, by=5))) %>% ggplot(aes(x = flags)) + geom_histogram(colour="black", fill="white", binwidth=0.5) + labs(title="Distribution of Flags By Age Bracket",x="Flags", y = "Frequency") + theme_minimal() + facet_wrap(~age_bracket, nrow = 4)

#By income_bracket
cust_dev_joined %>% mutate(income_bracket = cut(hhld_median_inc, breaks = seq(40000, 115000, by=5000), labels = c("(40000, 45000]", "(45000, 50000]", "(50000, 55000]", "(55000, 60000]", "(60000, 65000]", "(65000, 70000]", "(70000, 75000]", "(75000, 80000]", "(80000, 85000]", "(85000, 90000]", "(90000, 95000]", "(95000, 100000]", "(100000, 105000]", "(105000, 110000]", "(110000, 115000]"))) %>% ggplot(aes(x = flags)) + geom_histogram(colour="black", fill="white", binwidth=0.5) + labs(title="Distribution of Flags By Income Bracket",x="Flags", y = "Frequency") + theme_minimal()+ facet_wrap(~income_bracket, nrow = 3)
```

```{r}
#Plot with variables duration and flags to identify any correlation between them and other variables
#Emoji modifiers from B to F is from light to dark skin tone
cust_dev_joined %>% group_by(emoji_modifier) %>% ggplot(aes(x=duration, y=flags, colour = emoji_modifier)) + geom_point(size=0.1) + theme_minimal() + geom_smooth(method=lm, se=FALSE, fullrange=TRUE) + labs(title="Flags vs Duration, by Emoji Modifier",x="Duration (Minutes)", y = "Flags")

#Separated by emoji_modifer
cust_dev_joined %>% group_by(emoji_modifier) %>% ggplot(aes(x=duration, y=flags, colour = emoji_modifier)) + geom_point(size=0.1) + theme_minimal() + geom_smooth(method=lm, se=FALSE, fullrange=TRUE) + labs(title="Flags vs Duration, by Emoji Modifier",x="Duration (Minutes)", y = "Flags", colour = "Emoji Modifier") + facet_wrap(~emoji_modifier, nrow = 2)

#Separated by sex
cust_dev_joined %>% group_by(sex) %>% ggplot(aes(x=duration, y=flags, colour = sex)) + geom_point(size=0.1) + theme_minimal() + geom_smooth(method=lm, se=FALSE, fullrange=TRUE) +  labs(title="Flags vs Duration, by Sex",x="Duration (Minutes)", y = "Flags", colour = "Sex") + facet_wrap(~sex, nrow = 2)

#Separated by device_name
cust_dev_joined %>% group_by(device_name) %>% ggplot(aes(x=duration, y=flags, colour = device_name)) + geom_point(size=0.1) + theme_minimal() + geom_smooth(method=lm, se=FALSE, fullrange=TRUE) + labs(title="Flags vs Duration, by Device",x="Duration (Minutes)", y = "Flags", colour = "Device") + facet_wrap(~device_name, nrow = 4)

#Separated by age
cust_dev_joined %>% mutate(age_bracket = cut(age, breaks = seq(15, 95, by=5))) %>% group_by(age_bracket) %>% ggplot(aes(x=duration, y=flags, colour = age_bracket)) + geom_point(size=0.1) + theme_minimal() + geom_smooth(method=lm, se=FALSE, fullrange=TRUE) + labs(title="Flags vs Duration, by Age Bracket",x="Duration (Minutes)", y = "Flags", colour = "Age Bracket") + facet_wrap(~age_bracket, nrow = 4)

#Separated by income
cust_dev_joined %>% mutate(income_bracket = cut(hhld_median_inc, breaks = seq(40000, 115000, by=5000), labels = c("(40000, 45000]", "(45000, 50000]", "(50000, 55000]", "(55000, 60000]", "(60000, 65000]", "(65000, 70000]", "(70000, 75000]", "(75000, 80000]", "(80000, 85000]", "(85000, 90000]", "(90000, 95000]", "(95000, 100000]", "(100000, 105000]", "(105000, 110000]", "(110000, 115000]"))) %>% group_by(income_bracket) %>% ggplot(aes(x=duration, y=flags, colour = income_bracket)) + geom_point(size=0.1) + theme_minimal() + geom_smooth(method=lm, se=FALSE, fullrange=TRUE) + labs(title="Flags vs Duration, by Income Bracket",x="Duration (Minutes)", y = "Flags", colour = "Income Bracket") + facet_wrap(~income_bracket, nrow = 3)
```
```{r}
#Checking how age affects flags for different emoji_modifiers
cust_dev_joined %>% group_by(emoji_modifier) %>% ggplot(aes(x=age, y=flags, colour = emoji_modifier)) + geom_point(size=0.5) + theme_minimal() + geom_smooth(method=lm, se=FALSE, fullrange=TRUE) + labs(title="Flags vs Age, by Emoji Modifer",x="Age", y = "Flags", colour = "Emoji Modifier")

#Checking how income affects flags for different emoji_modifiers
cust_dev_joined %>% group_by(emoji_modifier) %>% ggplot(aes(x=hhld_median_inc, y=flags, colour = emoji_modifier)) + geom_point(size=0.5) + theme_minimal() + geom_smooth(method=lm, se=FALSE, fullrange=TRUE) + labs(title="Flags vs Household Median Income, by Emoji Modifier",x="Household Median Income", y = "Flags", colour = "Emoji Modifier")
```


```{r}
#Plotting log means of flags by age to check for linearity
#Interesting behaviour, weakly linear
cust_dev_joined %>% group_by(age) %>% summarise(flags=mean(flags)) %>% ggplot(aes(x=age, y=log(flags))) + geom_point() + geom_smooth() + theme_minimal() + labs(title="Log of Emperical Mean Flags by Age",x="Age", y = "log(Mean Flags)")

cust_dev_joined %>% group_by(age, emoji_modifier) %>% summarise(flags=mean(flags)) %>% ggplot(aes(x=age, y=log(flags), colour = emoji_modifier)) + geom_point() + theme_minimal() + geom_smooth() + labs(caption="Figure 2.10: Log of Emperical Mean Flags by Age, by Emoji Modifier",x="Age", y = "log(Mean Flags)", colour = "Emoji Modifier") + theme(plot.caption = element_text(size = 11))

#Plotting log means of flags by income to check for linearity
#Interesting behaviour, weakly linear
cust_dev_joined %>% group_by(hhld_median_inc) %>% summarise(flags=mean(flags)) %>% ggplot(aes(x=hhld_median_inc, y=log(flags))) + geom_point() + geom_smooth() + theme_minimal() + labs(title="Log of Emperical Mean Flags by Household Median Income",x="Household Median Income", y = "log(Mean Flags)")

cust_dev_joined %>% group_by(hhld_median_inc, emoji_modifier) %>% summarise(flags=mean(flags)) %>% ggplot(aes(x=hhld_median_inc, y=log(flags), colour = emoji_modifier)) + geom_point() + theme_minimal() + geom_smooth() + labs(caption="Figure 2.11: Log of Emperical Mean Flags by Household Median Income, by Emoji Modifier",x="Household Median Income", y = "log(Mean Flags)", colour = "Emoji Modifier") + theme(plot.caption = element_text(size = 11))
```

```{r}
#First attempt at a model
#After testing other models later in the code, this one is just better
model_base <- lme4::glmer(flags ~ factor(emoji_modifier) + (1|cust_id), family = poisson, offset = log(duration/60), data = cust_dev_joined)
summary(model_base)
lme4::VarCorr(model_base)
```

```{r}
#Other models to be considered

#0 coefficient models
model_0a <- lme4::glmer(flags ~ (1|cust_id), family = poisson, offset = log(duration/60), data = cust_dev_joined)
model_0b <- lme4::glmer(flags ~ (1|device_name), family = poisson, offset = log(duration/60), data = cust_dev_joined)
model_0c <- lme4::glmer(flags ~ (1|cust_id) + (1|device_name), family = poisson, offset = log(duration/60), data = cust_dev_joined)

#ANOVA shows model with only cust_id random intercept is the best
anova(model_0a, model_base, test = "Chisq")
anova(model_0b, model_base, test = "Chisq")
anova(model_0c, model_base, test = "Chisq")

#With device_name as random intercept
model_0d <- lme4::glmer(flags ~ factor(emoji_modifier) + (1|cust_id) + (1|device_name), family = poisson, offset = log(duration/60), data = cust_dev_joined)
#ANOVA shows device_name does not significantly improve the model
anova(model_0d, model_base, test = "Chisq")
```
```{r}
#1 coefficient models
model_null <- glm(flags ~ 1, family = poisson, offset = log(duration/60), data = cust_dev_joined)
model_1a <- glm(flags ~ age, family = poisson, offset = log(duration/60), data = cust_dev_joined)
model_1b <- glm(flags ~ hhld_median_inc, family = poisson, offset = log(duration/60), data = cust_dev_joined)
```
```{r}
#These are significant in the absence of our cust_id random intercept, but since we do have the intercept these effects are basically nested
#So ignore this chunk pretty much
anova(model_null, model_1a, test = "Chisq")
anova(model_null, model_1b, test = "Chisq")
```
```{r}
model_2a <- lme4::glmer(flags ~ factor(emoji_modifier) + factor(sex) + (1|cust_id), family = poisson, offset = log(duration/60), data = cust_dev_joined)

anova(model_base, model_2a)
```


```{r}
c_interval <- confint(model_base)
exp(c_interval)
```

```{r}
ests <- format(round(exp(summary(model_base)$coeff)[,1], 2), nsmall = 2)
cis <- format(round(exp(c_interval),2)[-1,], nsmall = 2)
conf_ints <- str_c("(", trimws(cis[,1]), ", ", cis[,2], ")")
rownames_for_table <- c("Baseline rate", "U+1F3FB", "U+1F3FC", "U+1F3FD", "U+1F3FE", "U+1F3FF")
colnames_for_table <- c("Estimate", "95% CI")
table <- cbind(ests, conf_ints)
rownames(table) <- rownames_for_table
colnames(table) <- colnames_for_table
knitr::kable(table, align = c("r", "r"))
```
```{r}
cust_dev_joined %>% group_by(emoji_modifier) %>% ggplot(aes(x=device_name, fill=emoji_modifier)) + geom_bar() + theme_minimal() + facet_wrap(~emoji_modifier, nrow = 3)
```

```{r}
res_int<- ranef(model_base)$cust_id$`(Intercept)`
qqnorm(res_int)
qqline(res_int)
plot(model_base)
```

